---
title: "NetworkHub: working with network data, seamlessly"
author: 
- name: Lotta Wagner
  affiliation: 
  - &id1 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: wagnerlo@uni-mainz.de
- name: Federico Marini
  affiliation: 
  - *id1
  email: marinif@uni-mainz.de
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    code_download: true
    global_numbering: true
    number_sections: true
editor_options: 
  chunk_output_type: console
link-citations: true
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('NetworkHub')`"
vignette: >
  %\VignetteIndexEntry{networkhub_string}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
---

# Installation 

```{r eval = FALSE}
if(!requireNamespace("NetworkHub", quietly = TRUE))
    install.packages("NetworkHub")
BiocManager::install("NetworkHub")
``` 


# Graph fetching

Initialization of the cache of "NetworkHub"

```{r init}
library("NetworkHub")
initialize_NetworkHub(nh_cachedir = "NetworkHub")
```

Extract the main core organisms of StringDB

```{r speciesinfo}
main_core_organisms <- info_species_stringdb() |> 
  dplyr::filter(STRING_type == "core") |> 
  head() |> 
  dplyr::pull(official_name_NCBI)

main_core_organisms 
```

Bringing the content into a table that later will be used to build a graph 

```{r getsdb}
sdb_human <- get_networkdata_stringdb("Homo sapiens",
                                      version = "12.0")
dim(sdb_human)
```

The lengthy part is to read in the accessory info, could be sped up with clever ways of reading in things I guess... #TODO: takes really long

```{r sdbprint}
head(sdb_human)
```


# Use case: DE on the graph!

Load in for example the results from differential expression #TODO: there is no package called ‘DeeDee’ 

```{r loadde}
# res object
data(res_de_macrophage, package = "GeneTonic")
res_de <- res_macrophage_IFNg_vs_naive
first_degenes <- na.omit(GeneTonic::deseqresult2df(res_de)[1:150, "id"])

library("macrophage")
library("DESeq2")
library("org.Hs.eg.db")
library("AnnotationDbi")

# dds object
data("gse", package = "macrophage")
dds_macrophage <- DESeqDataSet(gse, design = ~ line + condition)
rownames(dds_macrophage) <- substr(rownames(dds_macrophage), 1, 15)
dds_macrophage <- estimateSizeFactors(dds_macrophage)


#se_macrophage <- rbind(dds_macrophage, res_de) #TODO
    

#se_macrophage <- 
#  DeeDee::DeeDeeExperiment(
#    dds_macrophage, 
#    de_results = list(
#      macrores = res_de
#    )
#  )

#se_macrophage
```

DeeDee is just to put into the same SE object all the info on DE contrasts. That's a work in progress, but adding rowData info would also simply do it. #TODO

Going from the DE genes to the proteins...



```{r auxinfo}
aux_info_string <- get_accessoryinfo_stringdb(species = "Homo sapiens",
                                              version = "12.0")

anno_df_string <- create_annotation_from_stringdbaccessory(aux_info_string)

head(anno_df_string)

first_proteins <- anno_df_string$protein_id[match(first_degenes, anno_df_string$ensembl_id)] |> 
  na.omit()

first_proteins
```

Getting a subnetwork for the DE proteins...

```{r makegraph}
library("igraph")
de_proteins_subgraph <- build_graph_stringdb(graph_data = sdb_human, 
                                             output_format = "igraph",
                                             min_score_threshold = 200,
                                             subset_nodes = first_proteins)

plot(de_proteins_subgraph)

# re-sort the nodes alphabetically? makes sense
rank_gs <- rank(V(de_proteins_subgraph)$name)
de_proteins_subgraph <- permute.vertices(de_proteins_subgraph, rank_gs)
visNetwork::visIgraph(de_proteins_subgraph) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE),
             nodesIdSelection = TRUE)

V(de_proteins_subgraph)
str(V(de_proteins_subgraph))
```

Starting to have fun with interactive networks...  
Makes sense to display the gene names

```{r plotround1}
de_proteins_subgraph

de_graph <- de_proteins_subgraph

# renaming the nodes, using the gene symbols
V(de_graph)$name <- 
  anno_df_string$ensembl_id[match(V(de_graph)$name, anno_df_string$protein_id)]

de_graph <- igraph::induced.subgraph(de_graph, which(!is.na(V(de_graph)$name)))
de_graph_ens <- de_graph

# re-sort the nodes alphabetically? makes sense
rank_gs <- rank(V(de_graph)$name)
de_graph <- permute.vertices(de_graph, rank_gs)
visNetwork::visIgraph(de_graph) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE),
             nodesIdSelection = TRUE)
```

Now: putting on top the info from the expression values and their changes

```{r extraplots}
# different colors for the nodes in function of their logFC
# set.seed(42)
# fcs_genes <- rnorm(mean = 0, sd = 3, n = length(V(de_graph)))

str(V(de_graph))

intersect(V(de_graph)$name, first_degenes)

V(de_graph)$log2FC

V(de_graph)$log2FC <- res_de$log2FoldChange[match(V(de_graph)$name, rownames(res_de))]

mypal <- rev(scales::alpha(
  colorRampPalette(RColorBrewer::brewer.pal(name = "RdBu", 11))(50), 0.4
))

max_de <- max(abs(range(V(de_graph)$log2FC)))

V(de_graph)$color <- GeneTonic::map2color(V(de_graph)$log2FC, mypal, limits = c(-max_de, max_de))

gene_ids_backup <- V(de_graph)$name
gene_names <- anno_df_string$gene_symbol[match(gene_ids_backup, anno_df_string$ensembl_id)]
new_names <- gene_names
new_names[is.na(new_names)] <- gene_ids_backup[is.na(new_names)]
V(de_graph)$name <- new_names

rank_gs <- rank(V(de_graph)$name)
de_graph <- permute.vertices(de_graph, rank_gs)

# title for tooltips
V(de_graph)$title <- NA

V(de_graph)$title <- paste0(
  "<h4>", V(de_graph)$name, "</h4><br>",
  "logFC = ", format(round(V(de_graph)$log2FC, 2), nsmall = 2)
)

visNetwork::visIgraph(de_graph) |> 
  visNetwork::visEdges(color = list(color = "#88888888")) |> 
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                 degree = 1,
                                                 hover = TRUE),
                         nodesIdSelection = TRUE)
```


# Trying to go 1 on 1 vs the STRINGDB vignette!

```{r sdbvignette}
library("STRINGdb")
library("igraph")

string_db <- STRINGdb$new(version="11.5", 
                          species=9606,
                          score_threshold=200, 
                          network_type="full", 
                          input_directory="../dev/sdb_human/")

data(diff_exp_example1)
head(diff_exp_example1)

example1_mapped <- string_db$map( diff_exp_example1, "gene", removeUnmappedRows = TRUE )

hits <- example1_mapped$STRING_id[1:600]
string_db$plot_network(hits)

# get clusters
clustersList <- string_db$get_clusters(example1_mapped$STRING_id[1:600])
# plot first 4 clusters
par(mfrow=c(2,2))
for(i in seq(1:4)) {
  string_db$plot_network(clustersList[[i]])
}
```


```{r sdbnh}
# similarly, with our interactive thingoooooos...
ppi_human_v11.5 <- get_networkdata_STRINGDB(species = "Homo sapiens",
                                            version = "11.5")


network_desubset_600 <- build_graph_STRINGDB(graph_data = ppi_human_v11.5, 
                                             output_format = "igraph",
                                             min_score_threshold = 200,
                                             subset_nodes = example1_mapped$STRING_id[1:600])



# clustering the main graph like...

clu_graph_walktrap <- walktrap.community(network_desubset_600,merges=TRUE, modularity=TRUE)
clu_graph_fastgreedy <- fastgreedy.community(network_desubset_600,merges=TRUE, modularity=TRUE)
clu_graph_markov <- GeneTonic::cluster_markov(network_desubset_600)
clu_graph_louvain <- igraph::cluster_louvain(network_desubset_600)

table(membership(clu_graph_walktrap))
table(membership(clu_graph_fastgreedy))
table(membership(clu_graph_markov))
table(membership(clu_graph_louvain))

memb_to_use <- membership(clu_graph_walktrap)
memb_to_use <- membership(clu_graph_fastgreedy)
memb_to_use <- membership(clu_graph_louvain)

V(network_desubset_600)$membership <- memb_to_use
V(network_desubset_600)$color <- memb_to_use

visNetwork::visIgraph(network_desubset_600) %>%
   visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                    degree = 1,
                                                    hover = TRUE),
                            nodesIdSelection = TRUE,
                          selectedBy = "membership")
```


```{r allclusts}
clustersList_redone <- list()

for(i in 1:max(memb_to_use)){
  clustersList_redone[[i]] <- names(memb_to_use[memb_to_use == i])
}

cluster_networks <- lapply(clustersList_redone[1:4], function(arg) {
  build_graph_STRINGDB(graph_data = ppi_human_v11.5, 
                       output_format = "igraph",
                       min_score_threshold = 200,
                       subset_nodes = arg)
})

visNetwork::visIgraph(cluster_networks[[1]]) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                 degree = 1,
                                                 hover = TRUE),
                         nodesIdSelection = TRUE)

visNetwork::visIgraph(cluster_networks[[2]]) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                 degree = 1,
                                                 hover = TRUE),
                         nodesIdSelection = TRUE)

visNetwork::visIgraph(cluster_networks[[3]]) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                 degree = 1,
                                                 hover = TRUE),
                         nodesIdSelection = TRUE)

visNetwork::visIgraph(cluster_networks[[4]]) %>%
  visNetwork::visOptions(highlightNearest = list(enabled = TRUE,
                                                 degree = 1,
                                                 hover = TRUE),
                         nodesIdSelection = TRUE)

```


# Ok, bringing this into functions to make it streamlined

```{r mapse}
de_graph # would have to be with gene identifiers as in the SE
se_macrophage
annotation_df <- anno_df_string

rowdata_to_map <- "macrores_log2FoldChange"
assay_name <- "counts"


gg <- NetworkHub:::map_se_onto_graph(de_graph_ens, 
                                     se_macrophage, 
                                     anno_df_string, 
                                     assay_name = "counts")

gg

NetworkHub:::iplot_graph(gg, color_nodes_by = "macrores_log2FoldChange")

gg_genenames <- gg
genesymbols_matched <- anno_df_string$gene_symbol[match(V(gg)$name, anno_df_string$ensembl_id)]

scater::uniquifyFeatureNames(V(gg)$name, genesymbols_matched)
V(gg_genenames)$name <- scater::uniquifyFeatureNames(V(gg)$name, genesymbols_matched)

NetworkHub:::iplot_graph(gg_genenames, color_nodes_by = "macrores_log2FoldChange")
```



# Session Info {-}

```{r sessioninfo}
sessionInfo()
```

